### 2019-09-08 submission/manuscript.Rmd
Fix typos found while working on poster

**Omeprazole treatment did not promote susceptibility to *C. difficile infection* in mice.** Next, we examined whether omeprazole treatment altered susceptibility to *C. difficile* infection in mice. After omeprazole treatment or clindamycin treatment, mice were challenged with 10^3^ *C. difficile* 630 spores. While *C. difficile* colonized the clindamycin-treated mice, it did not colonize the omeprazole-treated mice (Figure 2A). Interestingly, only 1 cage of mice that received both omeprazole and clindamycin were colonized, while the other cage of mice were resistant (Figure 2A). The greatest shifts in bacterial communities occurred in the clindamycin-treated mice (Figure 2B, S2). Regardless of whether the mice became colonized, all of the mice had cleared *C. difficile* within 5 days (Figure 2A), suggesting that omeprazole did not affect the rate of clearance. Our results suggest that omeprazole treatment had no effect on bacterial community resistance to *C. difficile* colonization in mice. Instead most of the differences between the 3 treatment groups appeared to be driven by clindamycin administration (Figure 2C, S2). These findings demonstrated that high dose omeprazole treatment did not promote susceptibility to *C. difficile* colonization.

@@

Although several *C. difficile* mouse model studies have shown that PPIs have an effect on CDIs with or without additional antibiotic treatment [@Kaur2007; @kaur2011effect; @hung2015proton], there were insufficient controls to attribute the effect solely to PPI treatment. One group administered 0.5 mg/kg of the PPI lansoprazole daily for 2 weeks to mice and then challenged with *C. difficile* demonstrated that PPI treatment alone resulted in detectable *C. difficile* in the stool 1 week after challenge, however there was detectable *C. difficile* in mice not treated with antibiotics [@Kaur2007; @kaur2011effect]. The other mouse study demonstrated the antibiotic/PPI esomeprazole-treated mice developed more severe CDIs compared to antibiotic-treated mice, but the researchers did not have a group treated with just esomeprazole for comparison [@hung2015proton]. We tested the same high 40 mg/kg PPI dose and expanded pre-treatment to 7 days before challenge to test the impact of omeprazole treatment alone on our CDI mouse model. Additionally, we have previously demonstrated that mice from our breeding colony are resistant to *C. difficile* 630 colonization without antibiotic treatment [@Jenior2017], ensuring there was not already partial susceptiblity to *C. difficile* before treatment. The additional controls in our study allowed us to assess the contibution of omeprazole alone to *C. difficile* susceptibility in mice.  

@@

This research was supported by NIH grant U01AI12455. We would also like to thank the Unit for Laboratory Animal Medicine at the University of Michigan for maintaining our mouse colony and providing the infrastructure and support for performing our mouse expeirments. The authors are also thankful to members of the Schloss lab for helpful discussions throughout the process of designing the experiment, analyzing the results, crafting the figures, and drafting of the manuscript.

@@

**Animals.** All mouse experiments were performed with 7- to 12-week-old C57BL/6 male and female mice. Each experimental group of mice was spit between 2 cages with 2-3 mice housed per cage and male and female mice housed separately. All animal experiments were approved by the University of Michigan Animal Care and Use Committee (IACUC) under protocol number PRO00006983.

@@

**16S rRNA gene sequencing.** DNA for 16S rRNA gene sequencing was extracted from 10-50 mg fecal pellet from each mouse using the DNeasy Powersoil HTP 96 Kit (Qiagen) and an EpMotion 5075 automated pipetting system (Eppendorf). The 16S rRNA sequencing libary was prepared as described previously [@Kozich2013]. In brief, the ZymoBIOMICS^TM^ Microbial Community DNA Standard (Zymo, CA, USA) was used as a mock community [@Sze2019] and water was used as a negative control. The V4 hypervariable region of the 16S rRNA gene was amplified with Accuprime Pfx DNA polymerase (Thermo Fisher Scientific) using previously described custom barcoded primers [@Kozich2013]. The 16S rRNA amplicon library was sequenced with the MiSeq (Illumina). Amplicons were cleaned up and normalized with the SequalPrep Normalization Plate Kit (ThermoFisher Scientific) and pooled amplicons were quantified with the KAPA library quantification kit (KAPA Biosystems).

@@

**Statistical Analysis.** To test for differences in relative abundances in families and genera across our 3 different treatment groups at different timepoints (Clindamycin, Clind. + Omep., and Omeprazole on Day -7, 0, 2, and 9) or within the Omeprazole treatment group across 3 timepoints (Day -7, 0, and 9), we used a Kruskal-Wallis test with a Benjamini-Hochberg correction for multiple comparisons.

**Code availability.** The code for all sequence processing and analysis steps as well as an Rmarkdown version of this manuscript is available at https://github.com/SchlossLab/Tomkovich_PPI_mSphere_2019.

@@

## Figures
\includegraphics{figure_1.pdf}
**Figure 1. Omeprazole treatment had minimal impact on the murine fecal microbiota.** A. Mouse experiment timeline and logistics. The PPI omeprazole was administered throughout the duration of the experiment. Clindamycin was administered 1 day before *C. difficile* challenge on Day 0. Stools for 16S rRNA sequencing analysis were collected on the days that are labeled (Day -7, -5, -3, -1, 0, 1, 2, 3, 4, 5, 7, 9). *C. difficile* CFU in the stool was quantified daily through 6 days post-infection by anaerobic culture. B. Principal Coordinates Analaysis (PCoA) of Bray-curtis distances from stool samples of mice in the omeprazole treatment group during the intial 7 days of the experiment. Each color represents stool samples from the same mouse and lines connect sequentially collected samples. C-D. Relative abundances of families previously associated with PPI use in humans at the start of the experiment (C) and after 7 days of omeprazole treatment (D). Each circle represents an individual mouse. There were no significant differences acoss treatment groups for any of the identified families in the sequence data at day -7 (all P-values > 0.448) and day 0 (all P-values > 0.137), analyzed by Kruskal-Wallis test with a Benjamini-Hochberg correction for multiple comparisons. For C-D, the grey vertical line indicates the limit of detection.

\newpage
\includegraphics{figure_2.pdf}
**Figure 2. Omeprazole treatment alone does not promote CDIs in mice.** A. *C. difficile* CFUs/g stool measured each day post *C. difficile* challenge for clindamycin, clindamycin/omeprazole, and omeprazole-treated mice. Lines represent the mean CFU/g for each treatment group while points represent CFU/g for individual mice within each group. The black dashed line indicates the limit of detection. B. PCoA of of Bray-curtis distances from stool samples collected after antibiotic treatment (last 9 days of the experiment). Transparency of the symbol corresponds to treatment day. Symbols represent the *C. difficile* colonization status of the mice measured 2 days post-infection. Circles represent resistant mice (*C. difficile* was undetectable in stool samples), while X-shapes represent mice that were colonized with *C. difficile*, although all mice cleared *C. difficile* within 5 days of infection. Omeprazole treated fecal samples primarily cluster together throughout the experiment. C. Genera that vary the most across treatment groups for stool samples collected from mice 2 days post-infection. Data were analyzed by Kruskal-Wallis test, and no P-values were significant after Benjamini-Hochberg correction for multiple comparisons (all P-values > 0.092). The grey vertical line indicates the limit of detection.

\newpage
\includegraphics{figure_s1.pdf}
**Figure S1. Families within omeprazole treated mice fluctuate over time with no overall trend in either direction.** Relative abundance over time for *Lactobacillaceae* (A) and *Ruminococcaceae* (B), 2 of the PPI-associated familes from human PPI studies across all 3 treatment groups. Each point represents the relative abundance for an individual mouse stool sample, while the lines represent the mean relative abundances for each treatment group of mice. The grey horizontal lines indicate the limit of detection.

\newpage
\includegraphics{figure_s2.pdf}
**Figure S2. Microbiota diversity and richness decrease with antibiotic treatment but remain relatively constant with omeprazole treatment.** A. Boxplots of the Shannon Diversity Index values (A) and number of observed OTUs (B) for each group of mice over 3 timepoints (Day -7, 0, and 9). Each circle represents the value for a stool sample from an individual mouse.

>>>>

**Omeprazole treatment did not promote susceptibility to *C. difficile* infection in mice.** Next, we examined whether omeprazole treatment altered susceptibility to *C. difficile* infection in mice. After omeprazole treatment or clindamycin treatment, mice were challenged with 10^3^ *C. difficile* 630 spores. While *C. difficile* colonized the clindamycin-treated mice, it did not colonize the omeprazole-treated mice (Figure 2A). Interestingly, only 1 cage of mice that received both omeprazole and clindamycin were colonized, while the other cage of mice were resistant (Figure 2A). The greatest shifts in bacterial communities occurred in the clindamycin-treated mice (Figure 2B, S2). Regardless of whether the mice became colonized, all of the mice had cleared *C. difficile* within 5 days (Figure 2A), suggesting that omeprazole did not affect the rate of clearance. Our results suggest that omeprazole treatment had no effect on bacterial community resistance to *C. difficile* colonization in mice. Instead most of the differences between the 3 treatment groups appeared to be driven by clindamycin administration (Figure 2C, S2). These findings demonstrated that high dose omeprazole treatment did not promote susceptibility to *C. difficile* colonization.

@@

Although several *C. difficile* mouse model studies have shown that PPIs have an effect on CDIs with or without additional antibiotic treatment [@Kaur2007; @kaur2011effect; @hung2015proton], there were insufficient controls to attribute the effect solely to PPI treatment. One group administered 0.5 mg/kg of the PPI lansoprazole daily for 2 weeks to mice and then challenged with *C. difficile* demonstrated that PPI treatment alone resulted in detectable *C. difficile* in the stool 1 week after challenge, however there was detectable *C. difficile* in mice not treated with antibiotics [@Kaur2007; @kaur2011effect]. The other mouse study demonstrated the antibiotic/PPI esomeprazole-treated mice developed more severe CDIs compared to antibiotic-treated mice, but the researchers did not have a group treated with just esomeprazole for comparison [@hung2015proton]. We tested the same high 40 mg/kg PPI dose and expanded pre-treatment to 7 days before challenge to test the impact of omeprazole treatment alone on our CDI mouse model. Additionally, we have previously demonstrated that mice from our breeding colony are resistant to *C. difficile* 630 colonization without antibiotic treatment [@Jenior2017], ensuring there was not already partial susceptibility to *C. difficile* before treatment. The additional controls in our study allowed us to assess the contribution of omeprazole alone to *C. difficile* susceptibility in mice.  

@@

This research was supported by NIH grant U01AI12455. We would also like to thank the Unit for Laboratory Animal Medicine at the University of Michigan for maintaining our mouse colony and providing the infrastructure and support for performing our mouse experiments. The authors are also thankful to members of the Schloss lab for helpful discussions throughout the process of designing the experiment, analyzing the results, crafting the figures, and drafting of the manuscript.

@@

**Animals.** All mouse experiments were performed with 7- to 12-week-old C57BL/6 male and female mice. Each experimental group of mice was split between 2 cages with 2-3 mice housed per cage and male and female mice housed separately. All animal experiments were approved by the University of Michigan Animal Care and Use Committee (IACUC) under protocol number PRO00006983.

@@

**16S rRNA gene sequencing.** DNA for 16S rRNA gene sequencing was extracted from 10-50 mg fecal pellet from each mouse using the DNeasy Powersoil HTP 96 Kit (Qiagen) and an EpMotion 5075 automated pipetting system (Eppendorf). The 16S rRNA sequencing library was prepared as described previously [@Kozich2013]. In brief, the ZymoBIOMICS^TM^ Microbial Community DNA Standard (Zymo, CA, USA) was used as a mock community [@Sze2019] and water was used as a negative control. The V4 hypervariable region of the 16S rRNA gene was amplified with Accuprime Pfx DNA polymerase (Thermo Fisher Scientific) using previously described custom barcoded primers [@Kozich2013]. The 16S rRNA amplicon library was sequenced with the MiSeq (Illumina). Amplicons were cleaned up and normalized with the SequalPrep Normalization Plate Kit (ThermoFisher Scientific) and pooled amplicons were quantified with the KAPA library quantification kit (KAPA Biosystems).

@@

**Statistical Analysis.** To test for differences in relative abundances in families and genera across our 3 different treatment groups at different timepoints (Clindamycin, Clindamycin + Omeprazole, and Omeprazole on Day -7, 0, 2, and 9) or within the Omeprazole treatment group across 3 timepoints (Day -7, 0, and 9), we used a Kruskal-Wallis test with a Benjamini-Hochberg correction for multiple comparisons.

**Code availability.** The code for all sequence processing and analysis steps as well as a Rmarkdown version of this manuscript is available at https://github.com/SchlossLab/Tomkovich_PPI_mSphere_2019.

@@

## Figures
\includegraphics{figure_1.pdf}
**Figure 1. Omeprazole treatment had minimal impact on the murine fecal microbiota.** A. Mouse experiment timeline and logistics. The PPI omeprazole was administered throughout the duration of the experiment. Clindamycin was administered 1 day before *C. difficile* challenge on Day 0. Stools for 16S rRNA sequencing analysis were collected on the days that are labeled (Day -7, -5, -3, -1, 0, 1, 2, 3, 4, 5, 7, 9). *C. difficile* CFU in the stool was quantified daily through 6 days post-infection by anaerobic culture. B. Principal Coordinates Analysis (PCoA) of Bray-Curtis distances from stool samples of mice in the omeprazole treatment group during the initial 7 days of the experiment. Each color represents stool samples from the same mouse and lines connect sequentially collected samples. C-D. Relative abundances of families previously associated with PPI use in humans at the start of the experiment (C) and after 7 days of omeprazole treatment (D). Each circle represents an individual mouse. There were no significant differences across treatment groups for any of the identified families in the sequence data at day -7 (all P-values > 0.448) and day 0 (all P-values > 0.137), analyzed by Kruskal-Wallis test with a Benjamini-Hochberg correction for multiple comparisons. For C-D, the grey vertical line indicates the limit of detection.

\newpage
\includegraphics{figure_2.pdf}
**Figure 2. Omeprazole treatment alone does not promote CDIs in mice.** A. *C. difficile* CFUs/g stool measured each day post *C. difficile* challenge for clindamycin, clindamycin/omeprazole, and omeprazole-treated mice. Lines represent the mean CFU/g for each treatment group while points represent CFU/g for individual mice within each group. The black dashed line indicates the limit of detection. B. PCoA of of Bray-Curtis distances from stool samples collected after antibiotic treatment (last 9 days of the experiment). Transparency of the symbol corresponds to treatment day. Symbols represent the *C. difficile* colonization status of the mice measured 2 days post-infection. Circles represent resistant mice (*C. difficile* was undetectable in stool samples), while X-shapes represent mice that were colonized with *C. difficile*, although all mice cleared *C. difficile* within 5 days of infection. Omeprazole treated fecal samples primarily cluster together throughout the experiment. C. Genera that vary the most across treatment groups for stool samples collected from mice 2 days post-infection. Data were analyzed by Kruskal-Wallis test, and no P-values were significant after Benjamini-Hochberg correction for multiple comparisons (all P-values > 0.092). The grey vertical line indicates the limit of detection.

\newpage
\includegraphics{figure_s1.pdf}
**Figure S1. Families within omeprazole treated mice fluctuate over time with no overall trend in either direction.** Relative abundance over time for *Lactobacillaceae* (A) and *Ruminococcaceae* (B), 2 of the PPI-associated families from human PPI studies across all 3 treatment groups. Each point represents the relative abundance for an individual mouse stool sample, while the lines represent the mean relative abundances for each treatment group of mice. The grey horizontal lines indicate the limit of detection.

\newpage
\includegraphics{figure_s2.pdf}
**Figure S2. Microbiota diversity and richness decrease with antibiotic treatment but remain relatively constant with omeprazole treatment.** Boxplots of the Shannon Diversity Index values (A) and number of observed OTUs (B) for each group of mice over 3 timepoints (Day -7, 0, and 9). Each circle represents the value for a stool sample from an individual mouse.
